An Analysis of the Usage of the 2.4 GHz and 5 GHz Radio Bands in Antwerp

### 2017-05-27 advnetwpaper/paper_no_ieeetrans.tex
Some final changes to the report

Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license often use one of the unlicensed Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4 GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4 GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4 GHz band. In addition 802.11ac routers operating in the 5GHz range often use the 5GHz ISM band (which actually ranges from 5.725GHz to 5.875GHz). Some other bands in the 5GHz are also allowed for 802.11ac, although more restrictions, such as indoors-use only, apply there~\cite{50ism}. These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4 GHz and 5GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec\footnote{\url{https://www.imec-int.com/en/home}}. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On 11 of these gateways, we sampled activity on the 2.4 GHz and 5GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week. The main motivation of this experiment is to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.

@@

The ath10k version of spectral scan offers fewer options than the ath9k version. For instance the option to increase the scan time from \SI{4}{\micro\second} to \SI{2044}{\micro\second} is absent in ath10k. The only missing option relevant to our case is the \textit{chanscan} mode of spectral scan. When this mode is triggered the driver gathers a configurable number of samples for every channel it supports. We emulate this behaviour using the \textit{background} mode, which continuously scans the chipset's currently configured channel as long as it is not sending or receiving. We combine this with an \textit{iw scan} command, which listens for access point beacons for a provided list of frequencies. When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.

@@

Before going into the interpretation of this specific data, we provide a quick overview of the dBm unit, often used to express power ratios. The regular decibel (dB) unit is a dimensionless unit. In contrast dBm expresses absolute power in reference to watt. 0 dBm is defined as 1 milliwatt (mW). For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10. Milliwatt-to-dBm conversion is calculated using the formula\\ \\ $dBm = 10 log_{10}(\dfrac{x}{\SI{1}{\milli\watt}})$, with $x$ the power in mW.\\ \\ A negative dBm power simply means the power is less than \SI{1}{\milli\watt}. By using dBm, very small or large power values are often avoided. As power attenuates quadratically with distance, useful power values can vary dramatically.\\\\
For each channel scan, the driver reports the noise floor and RSSI value. The noise floor value is a hard-coded estimation for every channel and ranges between -94dBm and -108dBm. This value is the expected power of all noise in a channel combined. This includes noise sources such as thermal noise and cosmic noise. The chipset can only receive a signal whose power exceeds this noise floor. The RSSI value is a (unitless) integer indicating, as the name implies, the received power in the channel. While the existence of this value is part of the 802.11 specification, its calculation is not. As a result every manufacturer calculates the RSSI differently. Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI. These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm.~\cite{rssi-calc}. The RSSI reported by Atheros is often limited to 60. As RSSI was originally intended to estimate signal quality within the driver, values above 60 were not interesting; -36dBm (assuming a noise floor of -96dBm) already indicates excellent reception. It appears however that in this spectral scan mode, the hardware does return RSSI values of over 60. \\

@@

The earliest reference to this formula being used with the FOSS Atheros drivers was in an email on the ath9k-devel mailing list by Zefir Kurtisi~\cite{formula}, in reply to the Request For Comments (RFC) on spectral scan support by Simon W{\"u}nderlich. As far as we know neither this formula nor the calculation of its parameters were ever confirmed to be correct by Qualcomm Atheros. This remains an educated guess by the community.

@@

At first we measured the spectrum on every channel in both the 2.4 GHz and the 5GHz band every 10 minutes. The visualizations of the gathered data using Simon W{\"u}nderlich's FFT\_eval tool\footnote{https://github.com/simonwunderlich/FFT\_eval} showed that a shorter time between scans would be useful. There were often large variations between two subsequent samples, showing a use for a shorter interval. Additionally we noticed some extreme outliers. Occasionally a received signal strength of 129dBm or 8GW was reported. As a steady 8GW supply would require over 35 million solar panels~\cite{gigawatt}, so this is clearly an erroneous measurement. Although this was a fairly rare phenomenon, it still occurred up to a few times in each node's almost 10 million daily samples. To avoid such erroneous samples, we remove the $n$ highest signal strength measurements from every scan. The choice of $n$ depends on the sample size. For a single-channel 2.4 GHz scan, consisting of around 210 samples, we remove the top 5. We then consider the maximum signal strength among the remaining samples as the measured strength within the channel. \\
\begin{figure}[ht]
\begin{minipage}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{savitzky_0hours}
\caption{Savitzky-Golay smoothing.}
\label{savitzky_0}
\end{minipage}\
\begin{minipage}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{savitzky_1hour}
\caption{Savitzky-Golay smoothing, 1 additional hour of data point for each end.}
\label{savitzky_1}
\end{minipage}\
\end{figure}
\begin{figure}[htbp]
\begin{minipage}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{savitzky_4hours}
\caption{Savitzky-Golay smoothing, 4 additional hours of data points for each end.}
\label{savitzky_4}
\end{minipage}\
\end{figure}

We perform one scan per minute, alternating between 2.4 GHz and 5 GHz. As each scan takes at least 25 seconds to complete, we chose not to lower this interval any further. This generates around 860MB of raw data per node per day. Even when only considering the maximum signal strength per sample, we still need some way to condense this to a more digestible format. We experimented with some smoothing algorithms and had the best results with the Savitzky-Golay filter\footnote{\url{https://en.wikipedia.org/wiki/Savitzky\%E2\%80\%93Golay_filter}}. A detailed analysis of the filter was considered outside of the scope of this document, but we note that it is a digital filter intended for equally spaced data points that smooths data through convolution. We always used polynomials of degree 3. The only other parameter is the window size. The higher the window size, the smoother the final result. Figure \ref{savitzky_0} shows the resulting smoothings for 2 different window sizes. The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunchtime), while the larger window size is detailed enough to show that daytime is more active than nighttime. This filter performs well mainly for the center of the data set. The edges of the smoothing are clearly inaccurate, especially the sudden rise at the end of the day. This is a known issue with this filter. Luckily we usually also have the data of the preceding and following day. By starting the smoothing earlier and ending it later, we obtain a better smoothing. Figure \ref{savitzky_1} was generated using the data starting at 11PM the day before up to 1AM the day after. In figure \ref{savitzky_4} we increased this margin to 4 hours at each end. While the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours. We use this 4 hour buffer for all smoothings. Furthermore we decided to use the larger window size, the smoother graphs make it easier on the eyes to compare multiple smoothings on one graph. Additionally effects such as the drop during lunchtime were often not seen with tighter window sizes.\\
\newpage %Required to fix wrapfigure formatting
\begin{wrapfigure}{r}{0.35\textwidth}
\centering
\captionsetup{justification=centering}
\includegraphics[width=0.25\textwidth]{legend.png}
\caption{Legend:\\ days of the week}
\label{legend}
\end{wrapfigure}
Using an adapted version of Simon W{\"u}nderlich's FFT$\_$eval code, Python's Matplotlib library, the mathematical formulas mentioned above and a Savitzky-Golay filter with window size 361 we graphically represent our data points per node. For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days. For the 2.4 GHz band we perform this separately for the non-overlapping channels 1, 6 and 11. Each color in the plot represents a day of the week, using the legend in figure \ref{legend}. All other plots use the same legend, although days may be missing.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.75\textwidth]{50check.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{5 GHz spectral scan with active nearby device at 5.660GHz}
\label{fig_50}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.75\textwidth]{50check_nodata.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{5 GHz baseline spectral scan.}
\label{fig_50nodata}
\end{figure}
We first confirm that the spectral scan does in fact accurately measure activity per channel. To test this, we perform a data transfer with a laptop positioned roughly 50cm from the node. We perform this on available 2.4 GHz and 5 GHz networks. The 2.4 GHz network was using channel 6 (around 2.437GHz) and lead to the spectral scan in figure \ref{fig_24}. This scan shows a clear peak centered around channel 6 of around 20MHz wide. Figure \ref{fig_24nodata} shows a scan of the same node around the same time, but with our data transfer disabled. While the hazy peak around channel 1 is still there, the peak around channel 6 is not visible, indicating it was caused by our data transfer. Comparing the two, we see that our data transfer caused an increase in received signal strength of around 40dBm.\\ Our 5 GHz test was using an access point centered around channel 132 (5.66GHz) and lead to figure \ref{fig_50}. Compared to the flat \ref{fig_50nodata} captured with the data transfer disabled, the peak is again very clear. Received signal strength is increased by up to 50dBm in this case. The scan is detailed enough to indicate that the access point was using a 40MHz wide channel: data is being transferred both in channel 132 (5.66GHz) and channel 136 (5.68Ghz). These two tests clearly show that this method is accurate enough for our goals.

\section{Results}
In this section we present the results of our experiment. Due to (presumed) hardware failure, some measurements failed. As a result we had to limit and/or shift the timeframe of the presented measurements for some nodes. This is indicated below.
\subsection{2.4 GHz band}
\subsubsection{Recurring patterns between days} 
On the different nodes we had varying degrees of success in finding recurring patterns. In the 2.4 GHz band, two nodes that clearly showed repeating patterns across several days were node 8 at the Museum Plantin-Moretus and node 12 at the University of Antwerp, Middelheim campus. Both of these nodes detected patterns, with node 8 detecting a large increase in spectrum usage in the afternoon and node 12 detecting an increase during the daytime on weekdays. There were no peaks during the weekend. Node 8 was one of the nodes with failure, so we had to limit its timeframe.

@@

The most noticeable peak for node 12 is on channel 1 in figure \ref{node12-1}, with the received signal strength during the daytime being up to 20dBm higher than during the night time. During the weekend (black and yellow) there is no peak whatsoever. This indicates the traffic is most likely caused by Wi-Fi usage at the university, which is closed during the weekend. For the other channels there a noticeable peak during the daytime for some but not all weekdays. The weekend is still the quietest time of the week. Node 8 also shows a peak during the afternoon for channel 1 in \ref{node8-1}, except on Sunday. Surprisingly the museum is open on Sunday, but closed on Monday. Of the days with a peak, Monday's peak is the lowest. This seems to indicate that while visitors at the museum have some effect on the overall signal strength, the peak has another source, possibly an office nearby. \\ \\
We found no such peaks for any other nodes: spectrum usage was stable throughout the day. In the following section we investigate how active the different channels are.

@@

As shown in figures \ref{node3-1} to \ref{node9-11}, channel 1 is usually the busiest channel in the 2.4 GHz band, with channels 6 and 11 trailing far behind. Node 3 and 9, shown in those figures, measure a very busy medium around channel 1. However channels 6 and 11 are not particulary busy compared to the traffic seen on the other nodes, as we can see from these graphs:\\

All nodes covered so far were indoors nodes. We had access to one outdoors node, node 1, on which we also performed the experiment. On the indoors nodes we noticed that channel 6 and 11 were never any busier than channel 1. This is also true for node1: while activity is fairly constant throughout all tested days, channel 1 is again the busiest. Note that we again do not have a full week's data for this node. Note the remarkable difference between weekdays and the weekend here. For channel 11, the weekend is more quiet, however for channel 6 the weekend is the busiest period.

\subsection{5 GHz band}
\subsubsection{Recurring patterns between days}
\begin{figure}
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan1_image}
    \caption{Node 1, ch. 1, 13/05 - 17/05} \label{node1-1}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan6_image}
    \caption{Node 1, ch. 6, 13/05 - 17/05} \label{node1-6}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan11_image}
    \caption{Node 1, ch. 11, 13/05 - 17/05} \label{node1-11}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
\centering
\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
\caption{Node 3, ch. 36, 15/05 - 22/05}
\label{cot3_5ghz_pattern_chan36}
\end{minipage}\hfill
\end{figure}
The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements. Most of the channels we measured turned out to be silent or contain very little signal. This is partly to be expected. Many common communication devices only use the 2.4 GHz frequency band and many routers in use do not yet support the 5 GHz band. Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band. For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena. In most cases the received signal strength was very stable on all channels. We did notice some remarkable behaviour though. For example, node 3 measured a lower signal strength for some time around noon on both May 17\textsuperscript{th} and May 22\textsuperscript{nd} on channel 56, in figure \ref{cot3_5ghz_pattern_chan36}. The reason for this however is not as clear as with the daily peaks experienced in the 2.4 GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily. It could also simply be an anomaly. For this node, channel 36 was the only somewhat active channel. All other channels saw barely any traffic in the 5 GHz band. \\\\
It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes. The outdoor node has at least some activity on every channel whereas most of the indoors nodes show the bare minimum of activity for most channels. An example of the outdoors node's activity on channel 56 versus an indoors node on that same channel is shown in figures \ref{node1_5ghz_traffic_chan56} and \ref{node9_5ghz_traffic_chan56}. There could be some outside source of slight intermittent interference. As we only had access to one outdoors node we cannot determine whether this generally occurs with outside nodes, or is specific to this one.

@@

\subsubsection{Channel activity}
\begin{figure}[ht]
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan100_image.pdf}
	\caption{Node 3, ch. 120, 15/05 - 22/05}
	\label{node3_5ghz_traffic_chan100}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node8-student_2017-05-16_chan120_image.pdf}
	\caption{Node 8, ch. 120, 12/05 - 16/05}
	\label{node8_5ghz_traffic_chan120}
\end{minipage}\hfill
\end{figure}
\begin{figure}[ht]
\begin{minipage}{0.47\textwidth}
\centering
\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan140_image.pdf}
\caption{Node 1, ch. 140, 13/05 - 17/05}
\label{node1_5ghz_traffic_chan140}
\end{minipage}\hfill
\end{figure}
For all nodes except 1 (outdoor), 6 and 8, channels 100 and above seem to contain no activity. For the outside node this could merely be some outside interference, although more outside nodes are required to confirm this. For the other two nodes this might simply be a nearby 5 GHz router. Even for those nodes channels 100 and up are fairly silent.
Figures \ref{node3_5ghz_traffic_chan100} to \ref{node1_5ghz_traffic_chan140} show traffic on higher channels for nodes 3, 8 and 1. Each of these was somewhat busy on the 2.4 GHz band, and showed some activity on the lower 5 GHz channels. In the higher channels they barely exceed -80dBm. We do again notice shifts in signal strength with node 8, most likely caused by devices hopping to another frequency.

>>>>

Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license often use one of the unlicensed Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45 GHz, often called the 2.4 GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4 GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4 GHz band. In addition 802.11ac routers operating in the 5 GHz range often use the 5 GHz ISM band (which actually ranges from 5.725 GHz to 5.875 GHz). Some other bands in the 5 GHz are also allowed for 802.11ac, although more restrictions, such as indoors-use only, apply there~\cite{50ism}. These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4 GHz and 5 GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec\footnote{\url{https://www.imec-int.com/en/home}}. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On 11 of these gateways, we sampled activity on the 2.4 GHz and 5 GHz bands every 2 minutes for 12 days, including a Monday-to-Sunday week. The main motivation of this experiment is to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.

@@

The ath10k version of spectral scan offers fewer options than the ath9k version. For instance the option to increase the scan time from \SI{4}{\micro\second} to \SI{2044}{\micro\second} is absent in ath10k. The only missing option relevant to our case is the \textit{chanscan} mode of spectral scan. When this mode is triggered, the driver gathers a configurable number of samples for every channel it supports. We emulate this behaviour using the \textit{background} mode, which continuously scans the chipset's currently configured channel as long as it is not sending or receiving. We combine this with an \textit{iw scan} command, which listens for access point beacons for a provided list of frequencies. When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.

@@

Before going into the interpretation of this specific data, we provide a quick overview of the dBm unit, often used to express power ratios. The regular decibel (dB) unit is a dimensionless unit. The dBm unit instead expresses absolute power in reference to watt. 0dBm is defined as 1 milliwatt (mW). For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10. Milliwatt-to-dBm conversion is calculated using the formula\\ \\ $dBm = 10 log_{10}(\dfrac{x}{\SI{1}{\milli\watt}})$, with $x$ the power in mW.\\ \\ A negative dBm power simply means the power is less than \SI{1}{\milli\watt}. By using dBm, very small or large power values are often avoided. As power attenuates quadratically with distance, useful power values can vary dramatically.\\\\
For each channel scan, the driver reports the noise floor and RSSI value. The noise floor value is a hard-coded estimation for every channel and ranges between -94dBm and -108dBm. This value is the expected power of all noise in a channel combined. This includes noise sources such as thermal noise and cosmic noise. The chipset can only receive a signal whose power exceeds this noise floor. The RSSI value is a (unitless) integer indicating, as the name implies, the received power in the channel. While the existence of this value is part of the 802.11 specification, its calculation is not. As a result every manufacturer calculates the RSSI differently. Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI. As a result, the noise floor and RSSI are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm~\cite{rssi-calc}. The RSSI reported by Atheros is often limited to 60. As RSSI was originally intended to estimate signal quality within the driver, values above 60 were not interesting; -36dBm (assuming a noise floor of -96dBm) already indicates excellent reception. It appears however that in this spectral scan mode, the hardware does return RSSI values of over 60. \\

@@

The earliest reference to this formula being used with the FOSS Atheros drivers was in an email on the ath9k-devel mailing list by Zefir Kurtisi~\cite{formula}, in reply to the Request For Comments (RFC) on spectral scan support by Simon W{\"u}nderlich. As far as we know neither this formula nor the calculation of its parameters were ever confirmed to be correct by Qualcomm Atheros. This remains an educated guess by the community. \\ \\
For the remainder of this document, we will only take the noise floor and RSSI values into account. We saw no need for a granularity of the measurements lower than 20MHz.

@@

At first we measured the spectrum on every channel in both the 2.4 GHz and the 5 GHz band every 10 minutes. The visualizations of the gathered data using Simon W{\"u}nderlich's FFT\_eval tool\footnote{https://github.com/simonwunderlich/FFT\_eval} showed that a shorter time between scans would be useful. There were often large variations between two subsequent samples. Additionally we noticed some extreme outliers. Occasionally a received signal strength of 129dBm or 8GW was reported. As a steady 8GW supply would require over 35 million solar panels~\cite{gigawatt}, this is clearly an erroneous measurement. Although this was a fairly rare phenomenon, it still occurred up to a few times in each node's almost 10 million daily samples. To avoid such erroneous samples, we remove the $n$ highest signal strength measurements from every scan. The choice of $n$ depends on the sample size. For a single-channel 2.4 GHz scan, consisting of around 210 samples, we remove the top 5. We then consider the maximum signal strength among the remaining samples as the measured strength within the channel. \\
\begin{figure}[ht]
\begin{minipage}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{savitzky_0hours}
\caption{Savitzky-Golay smoothing.}
\label{savitzky_0}
\end{minipage}\
\begin{minipage}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{savitzky_1hour}
\caption{Savitzky-Golay smoothing, 1 additional hour of data point for each end.}
\label{savitzky_1}
\end{minipage}\
\end{figure}
\begin{figure}[htbp]
\begin{minipage}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{savitzky_4hours}
\caption{Savitzky-Golay smoothing, 4 additional hours of data points for each end.}
\label{savitzky_4}
\end{minipage}\
\end{figure}

We perform one scan per minute, alternating between 2.4 GHz and 5 GHz. As each scan takes at least 25 seconds to complete, we chose not to lower this interval any further. This generates around 860MB of raw data per node per day. Even when only considering the maximum signal strength per sample, we still need some way to condense this to a more digestible format. We experimented with some smoothing algorithms and had the best results with the Savitzky-Golay filter\footnote{\url{https://en.wikipedia.org/wiki/Savitzky\%E2\%80\%93Golay_filter}}. A detailed analysis of the filter was considered outside of the scope of this document, but we note that it is a digital filter intended for equally spaced data points that smooths data through convolution. We always used polynomials of degree 3. The only other parameter is the window size. The higher the window size, the smoother the final result. Figure \ref{savitzky_0} shows the resulting smoothings for 2 different window sizes. The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunchtime), while the larger window size is detailed enough to show that daytime is more active than nighttime. This filter performs well mainly for the center of the data set. The edges of the smoothing are clearly inaccurate, especially the sudden rise at the end of the day. This is a known issue with this filter. Luckily we usually also have the data of the preceding and following day. By starting the smoothing earlier and ending it later, we obtain a better smoothing. Figure \ref{savitzky_1} was generated using the data starting at 11PM the day before up to 1AM the day after. In figure \ref{savitzky_4} we increased this margin to 4 hours at each end. While the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours. We use this 4 hour buffer for all smoothings. Furthermore we decided to use the larger window size, as we noticed plots with multiple days combined were more readable with this size. Additionally effects such as the drop during lunchtime were often not seen even with tighter window sizes.\\
%\newpage %Required to fix wrapfigure formatting
\begin{wrapfigure}{r}{0.35\textwidth}
\centering
\captionsetup{justification=centering}
\includegraphics[width=0.25\textwidth]{legend.png}
\caption{Legend:\\ days of the week}
\label{legend}
\end{wrapfigure}
Using an adapted version of Simon W{\"u}nderlich's FFT$\_$eval code, Python's Matplotlib library, the mathematical formulas mentioned above and a Savitzky-Golay filter with window size 361 we graphically represent our data points per node. For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days. We repeat this for multiple channels in both bands. Each color in the plot represents a day of the week, using the legend in figure \ref{legend}. All other plots use the same legend, although days may be missing.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.75\textwidth]{50check.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{5 GHz spectral scan with active nearby device at 5.660GHz}
\label{fig_50}

\centering
\includegraphics[width=0.75\textwidth]{50check_nodata.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{5 GHz baseline spectral scan.}
\label{fig_50nodata}
\end{figure}
We first confirm that the spectral scan does in fact accurately measure activity per channel. To test this, we performed a data transfer with a laptop positioned roughly 50cm from the node, using available 2.4 GHz and 5 GHz networks. The 2.4 GHz network was using channel 6 (around 2.437GHz) and lead to the spectral scan in figure \ref{fig_24}. This scan shows a clear peak centered around channel 6 of around 20MHz wide. Figure \ref{fig_24nodata} shows a scan of the same node around the same time, but with our data transfer disabled. While the hazy peak around channel 1 is still there, the peak around channel 6 is not visible, indicating it was caused by our data transfer. Comparing the two, we see that our data transfer caused an increase in received signal strength of around 40dBm.\\ Our 5 GHz test was using an access point centered around channel 132 (5.66GHz) and lead to figure \ref{fig_50}. Compared to the flat figure \ref{fig_50nodata} captured with the data transfer disabled, the peak is again very clear. Received signal strength is increased by up to 50dBm in this case. The scan is detailed enough to indicate that the access point was using a 40MHz wide channel: data is being transferred both in channel 132 (5.66GHz) and channel 136 (5.68Ghz). These two tests clearly show that this method is accurate enough for our goals.

\section{Results}
In this section we present the results of our experiment. Due to (presumed) hardware failure, some measurements failed. As a result we had to limit and/or shift the timeframe of the presented measurements for some nodes. This is indicated below.
\subsection{2.4 GHz band}
\subsubsection{Recurring patterns between days} 
On the different nodes we had varying degrees of success in finding recurring patterns. In the 2.4 GHz band, two nodes that clearly showed repeating patterns across several days were node 8 at the Museum Plantin-Moretus and node 12 at the University of Antwerp, Middelheim campus. Both of these nodes detected patterns, with node 8 detecting a large increase in spectrum usage in the afternoon and node 12 detecting an increase during the daytime on weekdays. There were no peaks during the weekend. Node 8 was one of the failing nodes, so we had to limit its timeframe.

@@

The most noticeable peak for node 12 is on channel 1 in figure \ref{node12-1}, with the received signal strength during the daytime being up to 20dBm higher than during the night time. During the weekend (black and yellow) there is no peak whatsoever. This indicates the activity is most likely caused by Wi-Fi usage at the university, which is closed during the weekend. For the other channels there a noticeable peak during the daytime for some but not all weekdays. The weekend is still the quietest time of the week. Node 8 also shows a peak during the afternoon for channel 1 in \ref{node8-1}, except on Sunday. Surprisingly the museum is open on Sunday, but closed on Monday. Of the days with a peak, Monday's peak is the lowest. This seems to indicate that while visitors at the museum have some effect on the overall signal strength, the peak has another source, possibly an office nearby. \\ \\
We found no such peaks for any other nodes: spectrum usage was stable throughout the day. In the following section we investigate how active the different channels are.

@@

As shown in figures \ref{node3-1} to \ref{node9-11}, channel 1 is usually the busiest channel in the 2.4 GHz band, with channels 6 and 11 trailing far behind. Node 3 and 9, shown in those figures, measure a very busy medium around channel 1. However channels 6 and 11 are not particulary busy compared to other nodes. The two nodes show channel 1 activity of over -25dBm. This is likely caused by something else than a Wi-Fi router, although we are not sure what. This would likely make communication with the node over channel 1 very difficult.\\

All nodes covered so far were indoors nodes. We had access to one outdoors node, node 1, on which we also performed the experiment, as seen in figures \ref{node1-1} to \ref{node1-11}. On the indoors nodes we noticed that channel 6 and 11 were never any busier than channel 1. This is also true for node1: while activity is fairly constant throughout all tested days, channel 1 is again the busiest. Note that we again do not have a full week's data for this node. There is a remarkable difference between weekdays and the weekend here. For channel 11, the weekend is more quiet, however for channel 6 the weekend is the busiest period.

\subsection{5 GHz band}
\subsubsection{Recurring patterns between days}
\begin{figure}
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan1_image}
    \caption{Node 1, ch. 1, 13/05 - 17/05} \label{node1-1}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan6_image}
    \caption{Node 1, ch. 6, 13/05 - 17/05} \label{node1-6}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan11_image}
    \caption{Node 1, ch. 11, 13/05 - 17/05} \label{node1-11}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
\centering
\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
\caption{Node 3, ch. 36, 15/05 - 22/05}
\label{cot3_5ghz_pattern_chan36}
\end{minipage}\hfill
\end{figure}
The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements. Most of the channels we measured turned out to be silent or contain very little signal. This is partly to be expected. Many common communication devices only use the 2.4 GHz frequency band and many routers in use do not yet support the 5 GHz band. Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band. For the most part, the channels that did show activity did not show any significant patterns or recurring phenomena. In most cases the received signal strength was very stable on all channels. We did notice some remarkable behaviour though. For example, node 3 measured a lower signal strength for some time around noon on both May 17\textsuperscript{th} and May 22\textsuperscript{nd} on channel 56, in figure \ref{cot3_5ghz_pattern_chan36}. The reason for this however is not as clear as with the daily peaks experienced in the 2.4 GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily. It could also simply be an anomaly. For this node, channel 36 was the only somewhat active channel. All other channels saw barely any activity in the 5 GHz band. \\\\
It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes. The outdoor node has at least some activity on every channel whereas most of the indoors nodes show the bare minimum of activity for most channels. An example of the outdoors node's activity on channel 56 versus an indoors node on that same channel is shown in figures \ref{node1_5ghz_traffic_chan56} and \ref{node9_5ghz_traffic_chan56}. There could be some outside source of slight intermittent interference. As we only had access to one outdoors node we cannot determine whether this generally occurs with outside nodes, or is specific to this one.

@@

\subsubsection{Channel activity}
\begin{figure}[ht]
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan100_image.pdf}
	\caption{Node 3, ch. 120, 15/05 - 22/05}
	\label{node3_5ghz_traffic_chan100}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node8-student_2017-05-16_chan120_image.pdf}
	\caption{Node 8, ch. 120, 12/05 - 16/05}
	\label{node8_5ghz_traffic_chan120}
\end{minipage}\hfill
\end{figure}
\begin{figure}[ht]
\begin{minipage}{0.47\textwidth}
\centering
\includegraphics[width=\textwidth]{images/5_GHz/node1_2017-05-17_chan140_image.pdf}
\caption{Node 1, ch. 140, 13/05 - 17/05}
\label{node1_5ghz_traffic_chan140}
\end{minipage}\hfill
\end{figure}
For all nodes except 1 (outdoor), 6 and 8, channels 100 and above seem to contain no activity. For the outside node this could merely be some outside interference, although more outside nodes are required to confirm this. For the other two nodes this might simply be a nearby 5 GHz router. Even for those nodes channels 100 and up are fairly silent.
Figures \ref{node3_5ghz_traffic_chan100} to \ref{node1_5ghz_traffic_chan140} show activity on higher channels for nodes 3, 8 and 1. Each of these was somewhat busy on the 2.4 GHz band, and showed some activity on the lower 5 GHz channels. In the higher channels they barely exceed -80dBm. We do again notice shifts in signal strength with node 8, most likely caused by devices hopping to another frequency.
